---
title: "15-Side-by-side-gen-phen-comparison"
author: "Andrew Balmer"
date: "2024-03-11"
output: html_document
---

```{r setup, include=FALSE}
remove(list = ls())

library(ggExtra)
library(ggplot2)
library(ggpubr)
library(tidyverse)

options(dplyr.summarise.inform = FALSE)

source(file.path("..", "..", "tools", "notebook_helpers.R"))
analysis_paths <- amrc_notebook_setup(ensure_maps = TRUE)
generated_dir <- analysis_paths$generated_dir

tablemic_meta <- read.csv(
  file.path(generated_dir, "meta_data_Spneumoniae.csv"),
  header = TRUE,
  sep = ","
)

phen_env <- new.env(parent = emptyenv())
gen_env <- new.env(parent = emptyenv())
amrc_load_rdata(file.path(generated_dir, "Spneumo_3628_PCA_start_2D_METRIC.RData"), envir = phen_env)
amrc_load_rdata(file.path(generated_dir, "Spneumo_3628_PCA_start_2D_METRIC_genetic.RData"), envir = gen_env)

comparison_bundle <- amrc_prepare_spneumoniae_map_data(
  tablemic_meta = tablemic_meta,
  phenotype_mds = phen_env$torg_met,
  genotype_mds = gen_env$torg_met
)

cluster_12 <- amrc_cluster_map(
  data = comparison_bundle$pbp_data,
  coord_cols = c("G1", "G2"),
  n_clusters = 12,
  distinct_col = "PBP_type"
)

comparison_data <- amrc_add_cluster_assignments(
  data = comparison_bundle$data,
  assignments = cluster_12$assignments,
  key_col = "PBP_type",
  cluster_col = "gen_cluster"
)
comparison_data$gen_cluster <- as.factor(comparison_data$gen_cluster)

axis_spec <- function(values, padding = 1, step = 1) {
  lower <- floor(min(values, na.rm = TRUE) - padding)
  upper <- ceiling(max(values, na.rm = TRUE) + padding)
  list(
    limits = c(lower, upper),
    breaks = seq(lower, upper, by = step)
  )
}

phen_axis_x <- axis_spec(comparison_data$D1, padding = 1, step = 1)
phen_axis_y <- axis_spec(comparison_data$D2, padding = 1, step = 1)
gen_axis_x <- axis_spec(comparison_data$G1, padding = 5, step = 10)
gen_axis_y <- axis_spec(comparison_data$G2, padding = 5, step = 10)

label_adjustments <- data.frame(
  cluster = as.character(1:12),
  dx = c(-10, -8, -13, -9, -5, 0, 0, -8, -2, -15, -12, -10),
  dy = c(0, 9, 0, 5, -10, -9, 0, -5, 13, -10, 5, -10)
)

cluster_labels <- amrc_cluster_label_positions(
  data = comparison_data,
  x_col = "G1",
  y_col = "G2",
  cluster_col = "gen_cluster",
  adjustments = label_adjustments
)

gen_scale_bar <- list(
  x = min(comparison_data$G1, na.rm = TRUE) + 0.1 * diff(range(comparison_data$G1, na.rm = TRUE)),
  y = min(comparison_data$G2, na.rm = TRUE),
  length = 10,
  label = "10 AA",
  label_offset = 5,
  label_size = 4
)

genotype_plot <- amrc_plot_cluster_map(
  data = comparison_data,
  x = "G1",
  y = "G2",
  cluster_col = "gen_cluster",
  point_size = 3,
  limits_x = gen_axis_x$limits,
  breaks_x = gen_axis_x$breaks,
  limits_y = gen_axis_y$limits,
  breaks_y = gen_axis_y$breaks,
  scale_bar = gen_scale_bar
) +
  geom_label(
    data = cluster_labels,
    aes(x = x, y = y, label = cluster),
    inherit.aes = FALSE,
    size = 3.5,
    label.padding = grid::unit(0.15, "lines"),
    label.size = 0.25,
    fill = "white",
    color = "black"
  ) +
  theme(legend.position = "none")

genotype_plot_marginal <- ggExtra::ggMarginal(
  genotype_plot,
  size = 10,
  fill = "#E41A1C",
  alpha = 0.75,
  type = "histogram"
)

phenotype_plot <- amrc_plot_cluster_map(
  data = comparison_data,
  x = "D1",
  y = "D2",
  cluster_col = "gen_cluster",
  centroid_x = "x_centroid",
  centroid_y = "y_centroid",
  point_size = 1.5,
  centroid_size = 3,
  show_legend = TRUE,
  limits_x = phen_axis_x$limits,
  breaks_x = phen_axis_x$breaks,
  limits_y = phen_axis_y$limits,
  breaks_y = phen_axis_y$breaks
) +
  theme(legend.position = "none")

distance_table_samples <- amrc_compute_reference_distance_table(
  data = comparison_data,
  reference_pbp_type = "2-0-2",
  phenotype_cols = c("D1", "D2"),
  genotype_cols = c("G1", "G2"),
  cluster_col = "gen_cluster"
)

distance_summary_samples <- amrc_summarise_reference_distance_table(distance_table_samples)

distance_table_centroids <- amrc_compute_reference_distance_table(
  data = comparison_data,
  reference_pbp_type = "2-0-2",
  phenotype_cols = c("x_centroid", "y_centroid"),
  genotype_cols = c("G1", "G2"),
  phenotype_reference_cols = c("x_centroid", "y_centroid"),
  cluster_col = "gen_cluster"
)

distance_summary_centroids <- amrc_summarise_reference_distance_table(distance_table_centroids)

rho_label <- paste0(
  "rho = ",
  round(unname(distance_summary_centroids$correlation$estimate), 3),
  ", p-value < 0.001"
)

distance_relationship_plot <- amrc_plot_reference_distance_relationship(
  distance_table = distance_table_centroids,
  x_col = "gen_distance",
  y_col = "phen_distance",
  cluster_col = "gen_cluster",
  x_limits = c(0, 120),
  x_breaks = seq(0, 120, 10),
  y_limits = c(0, 16),
  y_breaks = seq(0, 16, 2),
  xlab = "Genetic distance (No. AA substitutions from 2-0-2)",
  ylab = "Median map distance (MIC units from 2-0-2)",
  annotation_text = rho_label,
  annotation_x = 60,
  annotation_y = 16
)
```

# Setup for plotting

The heavy data preparation for this notebook now happens in the package:

1. Load the canonical phenotype and genotype maps.
2. Calibrate them onto interpretable scales.
3. Build one comparison table with phenotype coordinates, genotype coordinates, metadata, and phenotype centroids.
4. Add genotype-map cluster assignments from the shared clustering helper.

That leaves this notebook focused on the actual comparison plots and summary tables.

```{r, echo = FALSE}
ggpubr::ggarrange(
  genotype_plot_marginal,
  phenotype_plot,
  nrow = 1,
  ncol = 2,
  font.label = list(size = 16, face = "bold")
)
```

## Distance from the 2-0-2 reference type

The first summary uses isolate-level phenotype and genotype positions. This matches the original table-building logic, but it now comes from one reusable helper rather than inline joins and manual centroid bookkeeping.

```{r, echo = FALSE}
distance_summary_samples$summary_with_overall
```

The second summary uses phenotype centroids for each PBP type. This is the version used for the phenotype-vs-genotype distance relationship plot below.

```{r, echo = FALSE}
distance_summary_centroids$summary_with_overall
summary(distance_summary_centroids$fit)
distance_summary_centroids$correlation
```

```{r, echo = FALSE}
distance_relationship_plot
```
