---
title: "Advanced Feature and Mixed-Model Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Feature and Mixed-Model Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(amrcartography)
```

## Why this vignette exists

The main package vignette focuses on the shortest generic workflow from a MIC
table to a map and optional external comparison. This vignette covers the more
advanced feature-analysis layer:

- gene presence/absence scans
- single-substitution style binary feature scans
- epistatic interaction markers
- mixed-model routes
- heritability and variance decomposition
- permutation scans
- kinship-BLUP style prediction

Most of the LIMIX-backed examples are shown with `execute = FALSE` so the
vignette stays lightweight and does not require a local Python/LIMIX
installation just to read the package documentation.

## A tiny generic fixture

```{r}
analysis_data <- data.frame(
  isolate_id = paste0("iso", 1:8),
  lineage = c("L1", "L1", "L1", "L2", "L2", "L3", "L3", "L3"),
  batch = c("B1", "B1", "B2", "B2", "B3", "B3", "B4", "B4"),
  map_x = c(0, 0.5, 1, 2, 2.5, 4, 4.5, 5),
  map_y = c(0, 0.4, 1, 2, 2.2, 4, 4.3, 5),
  drug_a = c(0.5, 1, 2, 2.5, 3, 4, 5, 5.5),
  drug_b = c(1, 1.5, 2.5, 2.8, 3.1, 4.2, 4.8, 5.1),
  gene_a = c(0, 0, 1, 0, 1, 1, 1, 1),
  gene_b = c(0, 1, 1, 0, 0, 1, 1, 1),
  gene_c = c(1, 0, 0, 1, 1, 0, 0, 1),
  stringsAsFactors = FALSE
)

kinship <- as.matrix(stats::dist(as.matrix(analysis_data[, c("map_x", "map_y")])))
rownames(kinship) <- analysis_data$isolate_id
colnames(kinship) <- analysis_data$isolate_id
```

## Simple gene or substitution scans

For fixed-effect scans against one or more responses, use the generic binary
feature scan helper:

```{r}
feature_scan <- amrc_scan_single_feature_associations(
  data = analysis_data,
  response_cols = c("drug_a", "drug_b"),
  feature_cols = c("gene_a", "gene_b"),
  covariate_cols = "lineage",
  min_group_size = 2
)

feature_scan$feature_summary
```

## Random-intercept mixed models

If you want a grouped random-intercept model directly in R:

```{r}
if (requireNamespace("lme4", quietly = TRUE)) {
  mixed_scan <- amrc_scan_single_feature_mixed_models(
    data = analysis_data,
    response_cols = c("drug_a", "drug_b"),
    feature_cols = c("gene_a", "gene_b"),
    random_effect_col = "batch",
    covariate_cols = "lineage",
    min_group_size = 2
  )

  mixed_scan$feature_summary
}
```

## Epistatic markers

The package can build pairwise interaction markers generically from binary
feature columns:

```{r}
epi_markers <- amrc_generate_epistatic_markers(
  data = analysis_data,
  feature_cols = c("gene_a", "gene_b", "gene_c"),
  id_col = "isolate_id"
)

epi_markers
```

## Optional LIMIX multivariate and epistatic scans

The manuscript-era multivariate mixed-model workflow is available as an
optional advanced route through the packaged Python/LIMIX script.

```{r eval = FALSE}
mv_scan <- amrc_run_limix_mvlmm(
  data = analysis_data,
  response_cols = c("drug_a", "drug_b"),
  marker_cols = c("gene_a", "gene_b", "gene_c"),
  covariate_cols = "lineage",
  id_col = "isolate_id",
  kinship_matrix = kinship,
  execute = TRUE
)

epi_scan <- amrc_run_limix_epistatic_scan(
  data = analysis_data,
  response_cols = c("drug_a", "drug_b"),
  feature_cols = c("gene_a", "gene_b", "gene_c"),
  covariate_cols = "lineage",
  id_col = "isolate_id",
  kinship_matrix = kinship,
  execute = TRUE
)
```

If the scan has already been executed and written result files, use the summary
helpers:

```{r eval = FALSE}
epi_summary <- amrc_summarise_limix_epistatic_scan(epi_scan)
epi_summary$summary
```

## Heritability and variance decomposition

The same optional LIMIX layer also exposes heritability estimation and
multi-component variance decomposition:

```{r eval = FALSE}
heritability <- amrc_run_limix_heritability(
  data = analysis_data,
  response_cols = c("map_x", "map_y"),
  kinship_matrix = kinship,
  id_col = "isolate_id",
  execute = TRUE
)

variance_parts <- amrc_run_limix_variance_decomposition(
  data = analysis_data,
  response_cols = c("map_x", "map_y"),
  kinship_components = list(
    component_a = kinship,
    component_b = kinship
  ),
  id_col = "isolate_id",
  execute = TRUE
)
```

## Permutation scans

For empirical null distributions, the package can stage or run response
permutations while keeping the marker matrix fixed.

```{r eval = FALSE}
perm_runs <- amrc_run_limix_permutation_scan(
  data = analysis_data,
  response_cols = c("map_x", "map_y"),
  marker_cols = c("gene_a", "gene_b", "gene_c"),
  covariate_cols = "lineage",
  id_col = "isolate_id",
  kinship_matrix = kinship,
  n_permutations = 10,
  execute = TRUE
)

perm_summary <- amrc_summarise_limix_permutation_scan(perm_runs)
perm_summary$permutation_summary
```

If you want the manuscript-style “sample a subset of interaction markers per
repeat” behavior, use `marker_sample_size`:

```{r eval = FALSE}
perm_runs <- amrc_run_limix_permutation_scan(
  data = analysis_data,
  response_cols = c("map_x", "map_y"),
  marker_cols = names(epi_markers)[-1],
  id_col = "isolate_id",
  n_permutations = 10,
  marker_sample_size = 2,
  execute = TRUE
)
```

## BLUP-style prediction

For a lighter-weight prediction workflow that stays in R:

```{r}
split <- amrc_make_train_test_split(nrow(analysis_data), proportion = 0.75, seed = 1)

blup_fit <- amrc_fit_kinship_blup(
  response = analysis_data$drug_a,
  kinship_matrix = kinship,
  train = split$train,
  test = split$test,
  lambda = 0.5,
  return_sd = TRUE
)

blup_fit$predictions
blup_fit$predictive_sd
```

Or use simple cross-validation:

```{r}
cv <- amrc_cross_validate_kinship_blup(
  response = analysis_data$drug_a,
  kinship_matrix = kinship,
  n_folds = 4,
  seed = 1,
  lambda = 0.5
)

cv$fold_summary
cv$overall
```

## Choosing among the options

- Use `amrc_scan_single_feature_associations()` for a quick generic scan.
- Use `amrc_scan_single_feature_mixed_models()` when you need a grouped random
  intercept and want to stay in R.
- Use the LIMIX wrappers when you specifically want the manuscript-style
  multivariate mixed-model route, permutation scans, heritability estimation,
  or multi-component variance decomposition.
- Use the BLUP helpers when your goal is prediction rather than association
  testing.
