---
title: "End-to-End S. pneumoniae Example Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-End S. pneumoniae Example Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows a small end-to-end workflow with `amrcartography`: start from bundled example inputs, preprocess phenotype and genotype data, fit phenotype and genotype maps, and build comparison summaries from the fitted maps.

The runnable example uses a tiny bundled raw-input example so the vignette stays fast and deterministic. A second non-evaluated chunk shows how to run the same workflow on the larger packaged *S. pneumoniae* example data.

## Load the package and locate the example inputs

```{r}
library(amrcartography)

mini_paths <- amrc_spneumoniae_example_paths("mini_raw")
mini_paths
```

## Preprocess phenotype and genotype inputs

```{r}
phenotype <- amrc_process_spneumoniae_phenotype(
  input_dir = mini_paths$root,
  save_outputs = FALSE
)

genotype <- amrc_process_spneumoniae_genotype(
  input_dir = mini_paths$root,
  metadata = phenotype$metadata,
  save_outputs = FALSE
)

phenotype$metadata
phenotype$mic
```

## Fit phenotype and genotype maps

```{r}
phenotype_fit <- amrc_compute_mds(
  phenotype$distance,
  itmax = 100,
  eps = 1e-06
)

genotype_fit <- amrc_compute_mds(
  genotype$distance,
  itmax = 100,
  eps = 1e-06
)

c(
  phenotype_stress = phenotype_fit$stress,
  genotype_stress = genotype_fit$stress
)
```

## Build a phenotype/genotype comparison table

```{r}
comparison_bundle <- amrc_prepare_spneumoniae_map_data(
  tablemic_meta = phenotype$metadata,
  phenotype_mds = phenotype_fit,
  genotype_mds = genotype_fit,
  phenotype_rotation_degrees = NULL,
  exclude_labids = character()
)

head(comparison_bundle$data)
```

## Cluster the genotype map and attach the cluster labels

```{r}
cluster_fit <- amrc_cluster_map(
  data = comparison_bundle$pbp_data,
  coord_cols = c("G1", "G2"),
  n_clusters = 2,
  distinct_col = "PBP_type",
  max_k = 3
)

comparison_data <- amrc_add_cluster_assignments(
  data = comparison_bundle$data,
  assignments = cluster_fit$assignments,
  key_col = "PBP_type",
  cluster_col = "gen_cluster"
)

cluster_fit$scree
comparison_data[, c("LABID", "PBP_type", "gen_cluster", "D1", "D2", "G1", "G2")]
```

```{r}
amrc_plot_cluster_elbow(cluster_fit$scree, highlight_cluster = 2, x_limits = c(1, 3))
```

```{r}
amrc_plot_cluster_map(
  data = comparison_data,
  x = "G1",
  y = "G2",
  cluster_col = "gen_cluster",
  point_size = 4,
  show_legend = TRUE
)
```

## Summarise phenotype vs genotype distance from a reference PBP type

```{r}
reference_distances <- amrc_compute_reference_distance_table(
  data = comparison_data,
  reference_pbp_type = comparison_data$PBP_type[[1]],
  cluster_col = "gen_cluster"
)

reference_summary <- amrc_summarise_reference_distance_table(
  distance_table = reference_distances,
  cluster_col = "gen_cluster"
)

reference_summary$summary_with_overall
```

```{r}
amrc_plot_reference_distance_relationship(
  reference_distances,
  cluster_col = "gen_cluster",
  xlab = "Genetic distance",
  ylab = "Phenotypic distance"
)
```

## Running the larger packaged example

The package also bundles processed example outputs and prebuilt canonical maps for the larger *S. pneumoniae* example used throughout this repository.

```{r}
generated_paths <- amrc_spneumoniae_example_paths("generated")

generated_paths[c(
  "mic_csv",
  "metadata_csv",
  "genotype_distance_rdata",
  "phenotype_map_metric",
  "genotype_map_metric"
)]
```

If you want to rebuild those larger example outputs from the raw example inputs, use the package pipeline helpers:

```{r, eval = FALSE}
out_dir <- tempfile("amrc-spneumoniae-generated-")

amrc_build_spneumoniae_example_outputs(
  raw_dir = tempfile("amrc-spneumoniae-raw-"),
  out_dir = out_dir,
  download_missing = TRUE
)

amrc_build_spneumoniae_example_maps(generated_dir = out_dir)
```

That larger workflow is the right starting point for reproducing the repository notebooks. The smaller workflow above is the better starting point for learning the package API.
