---
title: "Working with External Data Structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with External Data Structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the three generic ways to prepare the non-MIC structure
that `amrcartography` can compare against a phenotype map:

1. a precomputed external distance matrix
2. an aligned numeric feature table
3. an aligned character-state feature table

The example uses a tiny synthetic dataset so the workflow stays fast and easy
to adapt to your own data.

## Load the package

```{r}
library(amrcartography)
```

## Start from a small MIC table

```{r}
mic_table <- data.frame(
  isolate_id = paste0("iso", 1:5),
  lineage = c("L1", "L1", "L2", "L3", "L3"),
  drug_a = c(0.5, 1, 2, 4, 8),
  drug_b = c(1, 1, 2, 4, 8),
  drug_c = c(0.25, 0.5, 1, 2, 4),
  stringsAsFactors = FALSE
)

mic_data <- amrc_standardise_mic_data(
  data = mic_table,
  id_col = "isolate_id",
  mic_cols = c("drug_a", "drug_b", "drug_c"),
  metadata_cols = "lineage",
  transform = "log2"
)

phenotype_distance <- amrc_compute_mic_distance(mic_data)
phenotype_map <- amrc_compute_mds(phenotype_distance, itmax = 100, eps = 1e-06)
```

## Option 1: use a precomputed external distance matrix

If your external structure already exists as a distance matrix, standardise it
directly and fit the map from that object.

```{r}
precomputed_matrix <- matrix(
  c(
    0, 1, 2, 3, 4,
    1, 0, 2, 3, 4,
    2, 2, 0, 1, 2,
    3, 3, 1, 0, 1,
    4, 4, 2, 1, 0
  ),
  nrow = 5,
  byrow = TRUE,
  dimnames = list(mic_data$isolate_ids, mic_data$isolate_ids)
)

external_distance_precomputed <- amrc_compute_external_distance(precomputed_matrix)
as.matrix(external_distance_precomputed)
```

## Option 2: use an aligned numeric feature table

This is useful when your external measurements are already in a compact
coordinate or feature representation.

```{r}
numeric_features <- data.frame(
  isolate_id = c("iso3", "iso1", "iso2", "iso4", "iso5"),
  axis1 = c(1.1, 0.1, 0.2, 2.1, 2.4),
  axis2 = c(0.2, 0.0, 0.1, 1.8, 2.0),
  stringsAsFactors = FALSE
)

external_numeric <- amrc_standardise_external_data(
  data = numeric_features,
  id_col = "isolate_id",
  feature_mode = "numeric"
)

external_distance_numeric <- amrc_compute_external_feature_distance(external_numeric)
as.matrix(external_distance_numeric)
```

## Option 3: use aligned character-state features

This is useful when your external data are sequence- or allele-like aligned
states rather than numeric coordinates.

```{r}
character_features <- data.frame(
  isolate_id = c("iso1", "iso2", "iso3", "iso4", "iso5"),
  allele_1 = c("A", "A", "T", "T", "T"),
  allele_2 = c("C", "G", "G", "G", "C"),
  allele_3 = c("T", "T", "T", "C", "C"),
  stringsAsFactors = FALSE
)

external_character <- amrc_standardise_external_data(
  data = character_features,
  id_col = "isolate_id",
  feature_mode = "character"
)

external_distance_character <- amrc_compute_external_feature_distance(
  external_character,
  normalise_mismatch = TRUE
)

round(as.matrix(external_distance_character), 3)
```

## Build a phenotype-versus-external comparison table

Once you have an external distance object in any of those forms, the downstream
workflow is the same.

```{r}
external_map <- amrc_compute_mds(
  external_distance_numeric,
  itmax = 100,
  eps = 1e-06
)

comparison_bundle <- amrc_prepare_map_data(
  metadata = mic_data$metadata,
  phenotype_mds = phenotype_map,
  external_mds = external_map,
  id_col = "isolate_id",
  group_col = "lineage"
)

comparison_bundle$data[, c("isolate_id", "lineage", "D1", "D2", "E1", "E2")]
```

## Which input should you use?

- Use `amrc_compute_external_distance()` when you already trust the external
  distance matrix you want to analyse.
- Use `amrc_standardise_external_data()` plus
  `amrc_compute_external_feature_distance()` when you have aligned raw features
  and want the package to build the distance structure for you.
- Use numeric features when your external measurements are already in numeric
  form.
- Use character features when your external measurements are aligned states and
  mismatch distance is a sensible summary.
