---
title: "Public MIC Examples Across Species"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Public MIC Examples Across Species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows a small cross-species workflow using tiny public MIC
subsets bundled with the package. These are teaching/demo fixtures, not
benchmark datasets, but they are useful for checking that the generic workflow
still behaves sensibly across different bacterial species and different MIC
panels.

The bundled public examples are curated from public CDC & FDA Antimicrobial
Resistance Isolate Bank isolate-detail pages. The raw MIC values intentionally
retain censoring prefixes such as `<=` and `>`, so the generic MIC cleaning and
calibration path is exercised on real public values. Source panel links and the
recommended resource citation are collected in
`docs/PUBLIC_MIC_EXAMPLE_CITATIONS.md`.

## Load the package and inspect the bundled species examples

```{r}
if (!"package:amrcartography" %in% search()) {
  library(amrcartography)
}

public_specs <- amrc_public_mic_example_specs()
public_specs[, c("dataset_name", "species_group", "n_isolates", "suggested_mic_cols")]
```

The package currently includes six tiny public subsets spanning both Gram-
negative and Gram-positive organisms. This vignette keeps the worked example
to three species so it stays fast and compact.

## Define a lightweight species workflow helper

```{r}
build_species_example <- function(dataset_name) {
  spec <- public_specs[public_specs$dataset_name == dataset_name, , drop = FALSE]
  stopifnot(nrow(spec) == 1)

  mic_cols <- strsplit(spec$suggested_mic_cols, ",", fixed = TRUE)[[1]]
  data <- amrc_example_data(dataset_name)

  mic_data <- amrc_standardise_mic_data(
    data = data,
    id_col = "ar_bank_id",
    mic_cols = mic_cols,
    metadata_cols = c("species_group", "organism", "source", "biosample_accession"),
    transform = "log2",
    less_than = "numeric",
    greater_than = "numeric"
  )

  phenotype_distance <- amrc_compute_mic_distance(mic_data)
  phenotype_map <- amrc_compute_mds(phenotype_distance, itmax = 100, eps = 1e-06)
  calibrated <- amrc_calibrate_mds(phenotype_map)

  plot_data <- data.frame(
    ar_bank_id = rownames(calibrated$configuration),
    D1 = calibrated$configuration[, 1],
    D2 = calibrated$configuration[, 2],
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  plot_data <- merge(plot_data, mic_data$metadata, by = "ar_bank_id", all.x = TRUE, sort = FALSE)

  list(
    spec = spec,
    mic_cols = mic_cols,
    mic_data = mic_data,
    phenotype_distance = phenotype_distance,
    phenotype_map = phenotype_map,
    calibrated = calibrated,
    plot_data = plot_data,
    plot = amrc_plot_map(
      data = plot_data,
      x = "D1",
      y = "D2",
      point_fill = "#377EB8",
      grid_spacing = 1,
      use_cartography_theme = TRUE
    ) +
      ggplot2::labs(title = spec$species_group[[1]])
  )
}
```

## Build three small species-specific maps

```{r}
salmonella_example <- build_species_example("salmonella_enterica_mic")
campy_example <- build_species_example("campylobacter_jejuni_mic")
ecoli_example <- build_species_example("escherichia_coli_o157_mic")
```

Each species uses its own suggested MIC columns from the bundled manifest:

```{r}
list(
  salmonella = salmonella_example$mic_cols,
  campylobacter = campy_example$mic_cols,
  ecoli_o157 = ecoli_example$mic_cols
)
```

## Why calibration matters for 1-MIC spacing

The coordinates are calibrated before plotting. That is the intended route to
MIC-style map units. A one-unit grid should be interpreted as one doubling
dilution only after `amrc_fit_distance_calibration()` /
`amrc_calibrate_mds()` have been applied.

```{r}
rbind(
  data.frame(
    species = salmonella_example$spec$species_group,
    dilation = salmonella_example$calibrated$dilation
  ),
  data.frame(
    species = campy_example$spec$species_group,
    dilation = campy_example$calibrated$dilation
  ),
  data.frame(
    species = ecoli_example$spec$species_group,
    dilation = ecoli_example$calibrated$dilation
  )
)
```

## Compose the species maps with the manuscript-style panel helper

```{r}
if (requireNamespace("patchwork", quietly = TRUE)) {
  amrc_compose_manuscript_panel_grid(
    plots = list(
      salmonella_example$plot,
      campy_example$plot,
      ecoli_example$plot
    ),
    ncol = 3
  )
}
```

## Notes

- These are deliberately tiny public examples intended for documentation and
  validation.
- They are useful for testing the generic MIC cleaning and calibration flow on
  real public values.
- They are not large enough for any serious biological interpretation.
