---
title: "Using amrcartography with Your Own MIC Table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using amrcartography with Your Own MIC Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the generic package workflow: start from an arbitrary MIC
table, standardise it, compute a phenotype map, optionally align it to an
external distance structure, and then prepare reusable comparison, clustering,
and reference-distance summaries.

The example uses a small synthetic dataset so the workflow stays fast and easy
to adapt to your own data. The organism-specific *S. pneumoniae* case study is
available separately in `vignette("end-to-end-spneumoniae")`. If you want a
closer look at the different kinds of external structure the package can
accept, also see `vignette("external-data-structures")`.

## Load the package

```{r}
if (!"package:amrcartography" %in% search()) {
  library(amrcartography)
}
```

## Create a simple MIC table

The package expects one row per isolate, an isolate identifier column, and one
column per drug. Extra metadata columns can be carried through the workflow.

```{r}
mic_table <- data.frame(
  isolate_id = paste0("iso", 1:6),
  lineage = c("L1", "L1", "L2", "L2", "L3", "L3"),
  country = c("UK", "UK", "KE", "KE", "BR", "BR"),
  drug_a = c(0.5, 1, 2, 2, 4, 8),
  drug_b = c(1, 1, 2, 4, 4, 8),
  drug_c = c(0.25, 0.5, 1, 2, 4, 4),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

mic_table
```

## Standardise the MIC data

```{r}
mic_data <- amrc_standardise_mic_data(
  data = mic_table,
  id_col = "isolate_id",
  mic_cols = c("drug_a", "drug_b", "drug_c"),
  metadata_cols = c("lineage", "country"),
  transform = "log2"
)

mic_data$metadata
mic_data$mic
```

## Compute a phenotype distance matrix and phenotype map

```{r}
phenotype_distance <- amrc_compute_mic_distance(mic_data)

phenotype_map <- amrc_compute_mds(
  phenotype_distance,
  itmax = 100,
  eps = 1e-06
)

phenotype_report <- amrc_map_fit_report(phenotype_map)

phenotype_report$residual_summary
phenotype_report$stress_summary
```

## Add an optional external distance structure

Here we create a small synthetic external feature table, but in a real analysis
this could come from genotype-derived features, a phylogenetic embedding, or
any other aligned external measurements. If you already have a precomputed
external distance matrix, use `amrc_compute_external_distance()` directly. If
your external features are aligned character states, call
`amrc_standardise_external_data(feature_mode = "character")` and then
`amrc_compute_external_feature_distance()`.

```{r}
external_features <- data.frame(
  isolate_id = c("iso3", "iso1", "iso2", "iso4", "iso6", "iso5"),
  axis1 = c(0.0, 0.1, 0.3, 1.2, 2.2, 2.0),
  axis2 = c(0.0, 0.2, 0.1, 1.0, 2.1, 1.8),
  stringsAsFactors = FALSE
)

external_distance <- amrc_compute_external_feature_distance(
  data = external_features,
  id_col = "isolate_id"
)

external_map <- amrc_compute_mds(
  external_distance,
  itmax = 100,
  eps = 1e-06
)
```

## Build a reusable comparison table

```{r}
comparison_bundle <- amrc_prepare_map_data(
  metadata = mic_data$metadata,
  phenotype_mds = phenotype_map,
  external_mds = external_map,
  id_col = "isolate_id",
  group_col = "lineage"
)

comparison_bundle$data[, c(
  "isolate_id", "lineage", "country", "D1", "D2", "E1", "E2"
)]
```

Because `group_col = "lineage"` was supplied, the output also includes one row
per lineage plus phenotype-centroid columns.

```{r}
comparison_bundle$group_data[, c("lineage", "D1_centroid", "D2_centroid", "E1", "E2")]
```

## Cluster the external map and attach the labels

```{r}
cluster_fit <- amrc_cluster_map(
  data = comparison_bundle$group_data,
  coord_cols = c("E1", "E2"),
  distinct_col = "lineage",
  n_clusters = 2,
  max_k = 3
)

comparison_data <- amrc_add_cluster_assignments(
  data = comparison_bundle$data,
  assignments = cluster_fit$assignments,
  key_col = "lineage",
  cluster_col = "external_cluster"
)

comparison_data[, c("isolate_id", "lineage", "external_cluster", "D1", "D2", "E1", "E2")]
```

```{r}
amrc_plot_cluster_elbow(
  cluster_fit$scree,
  highlight_cluster = 2,
  x_limits = c(1, 3)
)
```

```{r}
amrc_plot_cluster_map(
  data = comparison_data,
  x = "E1",
  y = "E2",
  cluster_col = "external_cluster",
  point_size = 4,
  show_legend = TRUE
)
```

## Summarise distances from a reference lineage

```{r}
reference_distances <- amrc_compute_reference_distance_table(
  data = comparison_data,
  reference_value = "L1",
  reference_col = "lineage",
  id_col = "isolate_id",
  cluster_col = "external_cluster",
  phenotype_distance_col = "phenotype_distance",
  external_distance_col = "external_distance"
)

reference_summary <- amrc_summarise_reference_distance_table(
  distance_table = reference_distances,
  cluster_col = "external_cluster",
  phenotype_distance_col = "phenotype_distance",
  external_distance_col = "external_distance"
)

rbind(
  reference_summary$summary,
  reference_summary$average_row,
  reference_summary$sd_row
)
```

```{r}
amrc_plot_reference_distance_relationship(
  reference_distances,
  x_col = "external_distance",
  y_col = "phenotype_distance",
  cluster_col = "external_cluster",
  xlab = "External distance",
  ylab = "Phenotype distance"
)
```

## What to adapt for your own dataset

In a real analysis, the main things you would replace are:

- `mic_table`: your own isolate-by-drug MIC table
- `id_col`: your isolate identifier column
- `mic_cols`: the drug columns to include
- `metadata_cols`: the metadata you want to carry through the workflow
- `external_distance` or `external_features`: an optional aligned external
  structure

If you only need phenotype cartography, you can stop after
`amrc_compute_mic_distance()`, `amrc_compute_mds()`, and `amrc_map_fit_report()`.
