---
title: "Using amrcartography with Your Own MIC Table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using amrcartography with Your Own MIC Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the generic package workflow: start from an arbitrary MIC
table, standardise it, compute a phenotype map, optionally align it to an
external distance structure, and then prepare reusable comparison, clustering,
and reference-distance summaries.

The example uses a small synthetic dataset so the workflow stays fast and easy
to adapt to your own data. The organism-specific *S. pneumoniae* case study is
available separately in `vignette("end-to-end-spneumoniae")`. If you want a
closer look at the different kinds of external structure the package can
accept, also see `vignette("external-data-structures")`.

## Load the package

```{r}
if (!"package:amrcartography" %in% search()) {
  library(amrcartography)
}
```

## Load a simple MIC table

The package expects one row per isolate, an isolate identifier column, and one
column per drug. Extra metadata columns can be carried through the workflow.
This bundled example also includes common raw MIC censoring prefixes such as
`<`, `<=`, `>`, and `>=`. The generic parser also tolerates extra symbol clutter
around those values, for example `~0.5`, `-1`, or `<_0.25`.

```{r}
mic_table <- amrc_example_data("mic_raw")

mic_table
```

## Standardise the MIC data

By default the standardisation helpers strip censoring prefixes and keep the
reported numeric values. If your workflow prefers one dilution lower for `<`
values and one dilution higher for `>` values, use `less_than = "half"` and
`greater_than = "double"` as shown here.

```{r}
mic_data <- amrc_standardise_mic_data(
  data = mic_table,
  id_col = "isolate_id",
  mic_cols = c("drug_a", "drug_b", "drug_c"),
  metadata_cols = c("lineage", "source"),
  transform = "log2",
  less_than = "half",
  greater_than = "double"
)

mic_data$metadata
mic_data$mic
```

## Compute a phenotype distance matrix and phenotype map

```{r}
phenotype_distance <- amrc_compute_mic_distance(mic_data)

phenotype_map <- amrc_compute_mds(
  phenotype_distance,
  itmax = 100,
  eps = 1e-06
)

phenotype_report <- amrc_map_fit_report(phenotype_map)

phenotype_report$residual_summary
phenotype_report$stress_summary
```

## Add an optional external distance structure

Here we load a small generic numeric-feature table, but in a real analysis this
could come from genotype-derived features, a phylogenetic embedding, or any
other aligned external measurements. If you already have a precomputed external
distance matrix, use `amrc_compute_external_distance()` directly. If your
external features are aligned character states and you want a simple mismatch
distance, call
`amrc_standardise_external_data(feature_mode = "character")` and then
`amrc_compute_external_feature_distance()` or `amrc_compute_hamming_distance()`.
If the raw genotype/external features arrive as several tables, use
`amrc_bind_external_tables()` or `amrc_prepare_external_features()` first. If
you already have an aligned sequence or allele table and want the
`ape::dist.gene()` route used in the original genotype workflow, use
`amrc_compute_sequence_distance()` instead.

```{r}
external_features <- amrc_example_data("external_numeric")

external_distance <- amrc_compute_external_feature_distance(
  data = external_features,
  id_col = "isolate_id"
)

external_map <- amrc_compute_mds(
  external_distance,
  itmax = 100,
  eps = 1e-06
)
```

## Build a reusable comparison table

```{r}
comparison_bundle <- amrc_prepare_map_data(
  metadata = mic_data$metadata,
  phenotype_mds = phenotype_map,
  external_mds = external_map,
  id_col = "isolate_id",
  group_col = "lineage"
)

comparison_bundle$data[, c(
  "isolate_id", "lineage", "source", "D1", "D2", "E1", "E2"
)]
```

Because `group_col = "lineage"` was supplied, the output also includes one row
per lineage plus phenotype-centroid columns.

```{r}
comparison_bundle$group_data[, c("lineage", "D1_centroid", "D2_centroid", "E1", "E2")]
```

## Cluster the external map and attach the labels

```{r}
cluster_fit <- amrc_cluster_map(
  data = comparison_bundle$group_data,
  coord_cols = c("E1", "E2"),
  distinct_col = "lineage",
  n_clusters = 2,
  max_k = 3
)

comparison_data <- amrc_add_cluster_assignments(
  data = comparison_bundle$data,
  assignments = cluster_fit$assignments,
  key_col = "lineage",
  cluster_col = "external_cluster"
)

comparison_data[, c("isolate_id", "lineage", "external_cluster", "D1", "D2", "E1", "E2")]
```

```{r}
amrc_plot_cluster_elbow(
  cluster_fit$scree,
  highlight_cluster = 2,
  x_limits = c(1, 3)
)
```

```{r}
amrc_plot_cluster_map(
  data = comparison_data,
  x = "E1",
  y = "E2",
  cluster_col = "external_cluster",
  point_size = 4,
  show_legend = TRUE
)
```

## Summarise distances from a reference lineage

```{r}
reference_distances <- amrc_compute_reference_distance_table(
  data = comparison_data,
  reference_value = "L1",
  reference_col = "lineage",
  id_col = "isolate_id",
  cluster_col = "external_cluster",
  phenotype_distance_col = "phenotype_distance",
  external_distance_col = "external_distance"
)

reference_summary <- amrc_summarise_reference_distance_table(
  distance_table = reference_distances,
  cluster_col = "external_cluster",
  phenotype_distance_col = "phenotype_distance",
  external_distance_col = "external_distance"
)

rbind(
  reference_summary$summary,
  reference_summary$average_row,
  reference_summary$sd_row
)
```

```{r}
amrc_plot_reference_distance_relationship(
  reference_distances,
  x_col = "external_distance",
  y_col = "phenotype_distance",
  cluster_col = "external_cluster",
  xlab = "External distance",
  ylab = "Phenotype distance"
)
```

## Build a generic metadata-coloured map

Once you have a comparison table, you can also use the generic plotting
helpers directly. Here the phenotype map is coloured by lineage, overlaid with
one-unit grid spacing, and annotated with simple biplot-style arrows for the
three drug variables.

```{r}
map_plot <- amrc_plot_map(
  data = comparison_bundle$data,
  x = "D1",
  y = "D2",
  fill_col = "lineage",
  grid_spacing = 1,
  use_cartography_theme = TRUE,
  show_legend = TRUE
)

drug_vectors <- amrc_compute_biplot_vectors(
  data = cbind(comparison_bundle$data, mic_data$mic),
  x = "D1",
  y = "D2",
  variable_cols = c("drug_a", "drug_b", "drug_c")
)

amrc_add_biplot_vectors(map_plot, drug_vectors)
```

If you want calibrated tick-mark axes rather than simple arrows, use the
calibrated-axis helpers:

```{r}
drug_axes <- amrc_compute_calibrated_biplot_axes(
  data = cbind(comparison_bundle$data, mic_data$mic),
  x = "D1",
  y = "D2",
  variable_cols = c("drug_a", "drug_b")
)

amrc_add_calibrated_biplot_axes(map_plot, drug_axes)
```

The same generic plotting layer also supports faceting and ellipse-style group
envelopes.

```{r}
facet_plot <- amrc_plot_map(
  data = comparison_bundle$data,
  x = "D1",
  y = "D2",
  fill_col = "lineage",
  facet_by = "lineage",
  facet_ncol = 2
)

amrc_add_group_envelopes(
  plot = facet_plot,
  data = comparison_bundle$data,
  x = "D1",
  y = "D2",
  group_col = "lineage"
)
```

## Summarise distances between metadata groups

The generic comparison helpers can also summarise phenotype-versus-external
relationships at the level of arbitrary metadata groups.

```{r}
group_pairwise <- amrc_compute_group_pairwise_distances(
  data = comparison_bundle$data,
  group_col = "lineage",
  phenotype_cols = c("D1", "D2"),
  external_cols = c("E1", "E2"),
  summary_fun = "median"
)

group_pairwise
```

If you want sample-level within/between summaries rather than centroid-only
comparisons, use:

```{r}
group_distance_summary <- amrc_compute_group_distance_summary(
  data = comparison_bundle$data,
  group_col = "lineage",
  phenotype_cols = c("D1", "D2"),
  external_cols = c("E1", "E2")
)

group_distance_summary
```

## Scan binary genotype or gene features against phenotype responses

The same generic workflow can be used for simple gene presence/absence or
single-feature association scans.

```{r}
association_input <- cbind(
  comparison_bundle$data,
  mic_data$mic,
  has_gene_x = c(0, 0, 1, 1, 1, 1)
)

feature_scan <- amrc_scan_single_feature_associations(
  data = association_input,
  response_cols = c("drug_a", "drug_b", "drug_c"),
  feature_cols = "has_gene_x",
  min_group_size = 2
)

feature_scan$feature_summary
feature_scan$response_summary
```

If you want a true mixed-effects option with a random intercept, use the LMM
helpers. Here the random effect is just a toy grouping variable so the example
stays small.

```{r}
association_input$batch <- c("B1", "B1", "B2", "B2", "B3", "B3")
association_input$has_gene_y <- c(0, 1, 0, 1, 0, 1)

if (requireNamespace("lme4", quietly = TRUE)) {
  mixed_scan <- amrc_scan_single_feature_mixed_models(
    data = association_input,
    response_cols = c("drug_a", "drug_b", "drug_c"),
    feature_cols = "has_gene_x",
    random_effect_col = "batch",
    min_group_size = 2
  )

  mixed_scan$feature_summary
}
```

If you want the manuscript-style LIMIX mixed-model route, the package can also
stage and optionally run generic Python/LIMIX scans. This is the true
multivariate mixed-model option, but it requires a Python environment with
`limix`, `numpy`, and `pandas`, so the example below is shown without
execution.

```{r eval=FALSE}
limix_mv <- amrc_run_limix_mvlmm(
  data = association_input,
  response_cols = c("drug_a", "drug_b", "drug_c"),
  marker_cols = "has_gene_x",
  covariate_cols = "lineage",
  id_col = "isolate_id",
  execute = FALSE
)

limix_uv <- amrc_run_limix_lmm_scan(
  data = association_input,
  response_cols = c("drug_a", "drug_b", "drug_c"),
  marker_cols = "has_gene_x",
  covariate_cols = "lineage",
  id_col = "isolate_id",
  execute = FALSE
)

limix_mv$args
limix_uv$args
```

The same optional LIMIX layer also supports the other reusable pieces from the
old mixed-model workflow, again on generic inputs:

```{r eval=FALSE}
kinship <- as.matrix(stats::dist(external_features[, c("feature_1", "feature_2")]))
rownames(kinship) <- external_features$isolate_id
colnames(kinship) <- external_features$isolate_id

amrc_run_limix_heritability(
  data = association_input,
  response_cols = c("drug_a", "drug_b", "drug_c"),
  kinship_matrix = kinship,
  id_col = "isolate_id",
  execute = FALSE
)

amrc_run_limix_variance_decomposition(
  data = association_input,
  response_cols = c("drug_a", "drug_b", "drug_c"),
  kinship_components = list(component_a = kinship, component_b = kinship),
  id_col = "isolate_id",
  execute = FALSE
)

amrc_run_limix_epistatic_scan(
  data = association_input,
  response_cols = c("drug_a", "drug_b", "drug_c"),
  feature_cols = c("has_gene_x", "has_gene_y"),
  id_col = "isolate_id",
  execute = FALSE
)
```

For lighter-weight prediction workflows that do not require Python, the package
also includes generic kinship-BLUP helpers:

```{r}
toy_kinship <- as.matrix(stats::dist(as.matrix(external_features[, c("feature_1", "feature_2")])))

split <- amrc_make_train_test_split(nrow(association_input), proportion = 0.75, seed = 1)

blup_fit <- amrc_fit_kinship_blup(
  response = association_input$drug_a,
  kinship_matrix = toy_kinship,
  train = split$train,
  test = split$test,
  lambda = 0.5
)

blup_fit$predictions
```

## What to adapt for your own dataset

In a real analysis, the main things you would replace are:

- `mic_table`: your own isolate-by-drug MIC table
- `id_col`: your isolate identifier column
- `mic_cols`: the drug columns to include
- `metadata_cols`: the metadata you want to carry through the workflow
- `less_than` and `greater_than`: how to interpret censored MIC strings before
  numeric coercion and log transformation
- `external_distance` or `external_features`: an optional aligned external
  structure

If you only need phenotype cartography, you can stop after
`amrc_compute_mic_distance()`, `amrc_compute_mds()`, and `amrc_map_fit_report()`.
