---
title: "Chapter 3 - S. pneumoniae - Clustering and informative isolates (choice of clustering for genotype map)"
author: "Andrew Balmer"
date: "2021"
output: html_document
---

```{r setup, include=FALSE}
remove(list = ls())

library(ggExtra)
library(ggplot2)
library(cluster)
library(tidyverse)

options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(message = FALSE)

source(file.path("..", "..", "tools", "notebook_helpers.R"))
analysis_paths <- amrc_notebook_setup(ensure_maps = TRUE)
generated_dir <- analysis_paths$generated_dir

tablemic_meta <- read.csv(
  file.path(generated_dir, "meta_data_Spneumoniae.csv"),
  header = TRUE,
  sep = ","
)

phen_env <- new.env(parent = emptyenv())
gen_env <- new.env(parent = emptyenv())
amrc_load_rdata(file.path(generated_dir, "Spneumo_3628_PCA_start_2D_METRIC.RData"), envir = phen_env)
amrc_load_rdata(file.path(generated_dir, "Spneumo_3628_PCA_start_2D_METRIC_genetic.RData"), envir = gen_env)

comparison_bundle <- amrc_prepare_spneumoniae_map_data(
  tablemic_meta = tablemic_meta,
  phenotype_mds = phen_env$torg_met,
  genotype_mds = gen_env$torg_met
)

cluster_12 <- amrc_cluster_map(
  data = comparison_bundle$pbp_data,
  coord_cols = c("G1", "G2"),
  n_clusters = 12,
  distinct_col = "PBP_type"
)

clustered_12 <- amrc_add_cluster_assignments(
  data = comparison_bundle$data,
  assignments = cluster_12$assignments,
  key_col = "PBP_type",
  cluster_col = "gen_cluster"
)
clustered_12$gen_cluster <- as.factor(clustered_12$gen_cluster)

cluster_11 <- amrc_cluster_map(
  data = comparison_bundle$pbp_data,
  coord_cols = c("G1", "G2"),
  n_clusters = 11,
  distinct_col = "PBP_type"
)

clustered_11 <- amrc_add_cluster_assignments(
  data = clustered_12,
  assignments = cluster_11$assignments,
  key_col = "PBP_type",
  cluster_col = "gen_cluster_11"
)
clustered_11$gen_cluster_11 <- as.factor(clustered_11$gen_cluster_11)

axis_spec <- function(values, padding = 1, step = 1) {
  lower <- floor(min(values, na.rm = TRUE) - padding)
  upper <- ceiling(max(values, na.rm = TRUE) + padding)
  list(
    limits = c(lower, upper),
    breaks = seq(lower, upper, by = step)
  )
}

phen_axis_x <- axis_spec(clustered_12$D1, padding = 1, step = 1)
phen_axis_y <- axis_spec(clustered_12$D2, padding = 1, step = 1)
gen_axis_x <- axis_spec(clustered_12$G1, padding = 5, step = 10)
gen_axis_y <- axis_spec(clustered_12$G2, padding = 5, step = 10)

gen_scale_bar <- list(
  x = min(clustered_12$G1, na.rm = TRUE) + 0.1 * diff(range(clustered_12$G1, na.rm = TRUE)),
  y = min(clustered_12$G2, na.rm = TRUE),
  length = 10,
  label = "10 AA",
  label_offset = 5,
  label_size = 5
)

genotype_plot_12 <- amrc_plot_cluster_map(
  data = clustered_12,
  x = "G1",
  y = "G2",
  cluster_col = "gen_cluster",
  point_size = 4.5,
  limits_x = gen_axis_x$limits,
  breaks_x = gen_axis_x$breaks,
  limits_y = gen_axis_y$limits,
  breaks_y = gen_axis_y$breaks,
  scale_bar = gen_scale_bar
)

phenotype_plot_12 <- amrc_plot_cluster_map(
  data = clustered_12,
  x = "D1",
  y = "D2",
  cluster_col = "gen_cluster",
  background_data = clustered_12,
  centroid_x = "x_centroid",
  centroid_y = "y_centroid",
  point_size = 2.5,
  centroid_size = 3.25,
  facet_by = "gen_cluster",
  limits_x = phen_axis_x$limits,
  breaks_x = phen_axis_x$breaks,
  limits_y = phen_axis_y$limits,
  breaks_y = phen_axis_y$breaks
)

separation_all <- amrc_summarise_cluster_separation(
  data = clustered_12,
  coord_cols = c("D1", "D2"),
  cluster_col = "gen_cluster"
)
separation_distinct <- amrc_summarise_cluster_separation(
  data = clustered_12,
  coord_cols = c("D1", "D2"),
  cluster_col = "gen_cluster",
  distinct_col = "PBP_type"
)

genotype_plot_11 <- amrc_plot_cluster_map(
  data = clustered_11,
  x = "G1",
  y = "G2",
  cluster_col = "gen_cluster_11",
  point_size = 4,
  point_alpha = 0.6,
  show_legend = TRUE,
  limits_x = gen_axis_x$limits,
  breaks_x = gen_axis_x$breaks,
  limits_y = gen_axis_y$limits,
  breaks_y = gen_axis_y$breaks
)

pam_input <- clustered_12 %>%
  distinct(PBP_type, .keep_all = TRUE) %>%
  select(G1, G2)

pam_fit <- cluster::pam(pam_input, metric = "euclidean", k = 11)
pam_plot_data <- cbind(pam_input, pam_cluster = as.factor(pam_fit$clustering))

pam_plot <- amrc_plot_cluster_map(
  data = pam_plot_data,
  x = "G1",
  y = "G2",
  cluster_col = "pam_cluster",
  point_size = 4.5,
  point_alpha = 0.6,
  limits_x = gen_axis_x$limits,
  breaks_x = gen_axis_x$breaks,
  limits_y = gen_axis_y$limits,
  breaks_y = gen_axis_y$breaks
)
```

## Clustering on genotype map

In total, there were 305 PBP types in the collection of 3628 isolates with all 6 drugs measured. This notebook now uses the package comparison helpers to build the phenotype/genotype joint table, cluster the genotype map, and summarise how well those genotype clusters separate positions on the phenotype map.

### Choosing the optimal number of clusters

The package workflow here follows the same logic as the original notebook:

1. Build a calibrated phenotype/genotype comparison table.
2. Cluster the genotype map using Ward hierarchical clustering.
3. Examine the dendrogram and the elbow-style within-cluster inertia curve.
4. Compare 12 clusters with the nearby 11-cluster solution.
5. Check how well genotype-defined clusters separate phenotype positions.

### Dendrogram method

```{r, echo = FALSE}
plot(cluster_12$hclust, labels = FALSE, hang = -1, main = "Ward hierarchical clustering on genotype map")
rect.hclust(cluster_12$hclust, k = 12, border = "red")
```

### Elbow method

```{r, echo = FALSE}
amrc_plot_cluster_elbow(cluster_12$scree, highlight_cluster = 3, x_limits = c(1, 20))
```

```{r, echo = FALSE}
amrc_plot_cluster_elbow(
  cluster_12$scree,
  highlight_cluster = 12,
  x_limits = c(2, 20),
  y_limits = c(0, max(cluster_12$scree$within_cluster_inertia[cluster_12$scree$n_clusters >= 2]) * 0.35),
  draw_path = TRUE
)
```

### Clustered genotype and phenotype maps

```{r, echo = FALSE}
ggExtra::ggMarginal(genotype_plot_12, size = 10, fill = "#E41A1C", alpha = 0.9, type = "histogram")
```

```{r, echo = FALSE}
phenotype_plot_12 +
  theme(strip.background = element_blank(), strip.text.x = element_blank())
```

### Separation of phenotype space by genotype cluster

```{r, echo = FALSE}
amrc_plot_distance_histogram(separation_all$hist_data)
separation_all$summary
```

The first separation summary uses all isolates. The second repeats the same calculation after collapsing to one row per unique PBP type.

```{r, echo = FALSE}
amrc_plot_distance_histogram(separation_distinct$hist_data)
separation_distinct$summary
```

### Using 11 clusters

I re-ran the same clustering with 11 groups so the nearby solution can be compared directly with the 12-cluster choice used elsewhere in the analysis.

```{r, echo = FALSE}
genotype_plot_11
```

### Comparison of 11/12 clusters

Comparing 11 versus 12 clusters still gives a very similar partitioning. The main change is that one broad region on the left side of the genotype map is split more finely under the 12-cluster solution, which is why 12 remained the more interpretable choice in the downstream analyses.

```{r, echo = FALSE}
genotype_plot_12
genotype_plot_11
```

### Comparison with K-medoids clustering algorithm

The original notebook also compared hierarchical clustering with K-medoids. That comparison is kept here, but the plot is now built from the canonical package-prepared genotype coordinates rather than from notebook-specific data wrangling.

```{r, echo = FALSE}
pam_plot
```
