---
title: "11-Dimensionality-testing-MDS-plot"
author: "Andrew Balmer"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
remove(list = ls())
set.seed(123)

library(ggplot2)
library(tidyverse)

source(file.path("..", "..", "tools", "notebook_helpers.R"))
analysis_paths <- amrc_notebook_setup(ensure_maps = TRUE)
generated_dir <- analysis_paths$generated_dir

tablemic <- read.csv(
  file.path(generated_dir, "MIC_table_Spneumoniae.csv"),
  header = TRUE,
  sep = ","
)
tablemic_meta <- read.csv(
  file.path(generated_dir, "meta_data_Spneumoniae.csv"),
  header = TRUE,
  sep = ","
)

phen_env <- new.env(parent = emptyenv())
amrc_load_rdata(file.path(generated_dir, "Spneumo_3628_PCA_start_2D_METRIC.RData"), envir = phen_env)
torg_met <- phen_env$torg_met

shift_positive <- function(df) {
  out <- df
  for (i in seq_len(ncol(out))) {
    out[, i] <- out[, i] + (-(min(out[, i], na.rm = TRUE))) + 1
  }
  out
}

tablemic_shifted <- shift_positive(tablemic)

single_column_permutation <- amrc_single_column_permutation_study(
  reference_fit = torg_met,
  tablemic = tablemic_shifted,
  nrep = 100,
  seed = 123
)

one_dimensional_noise <- amrc_one_dimensional_noise_study(
  tablemic = tablemic_shifted,
  nrep = 100,
  noise_pct = 10,
  seed = 123,
  mds_args = list(modulus = 1, itmax = 1000, eps = 1e-06)
)

one_vs_two_dimensional <- amrc_compare_one_and_two_dimensional_maps(
  one_dim_fit = one_dimensional_noise$observed_fit,
  two_dim_fit = torg_met,
  lab_ids = tablemic_meta$LABID,
  rotation_degrees = 326
)
```

## Phenotypic covariation in beta-lactam MIC is low, but not one-dimensional

The original notebook tested two related hypotheses:

1. Could the apparent MIC covariation arise under random permutation of the observed values?
2. Could the phenotype map be mostly one-dimensional, with the second dimension mainly reflecting assay noise?

Those analyses are now implemented with package functions, so this notebook mostly just runs them and plots the results.

### Permuting a single column at a time

For each dataset, the MIC values are permuted while preserving the marginal value distribution. If the observed map stress were typical under those permutations, that would suggest the map structure could arise by chance.

```{r, echo = FALSE}
amrc_plot_histogram_with_reference(
  values = single_column_permutation$stress,
  xlab = "Stress",
  fill = "#E41A1C",
  bins = 50,
  mean_line = TRUE,
  reference_line = torg_met$stress
) +
  coord_fixed(ratio = 0.0005) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16)
  )
```

```{r, echo = FALSE}
single_column_permutation$summary %>%
  mutate(across(everything(), round, 3))
```

## One-dimensional data with added noise

The second analysis creates synthetic one-dimensional MIC tables by copying one drug column across all drugs and perturbing 10% of the values by one log2 dilution in either direction. The stress of those synthetic one-dimensional datasets is then compared with the stress of a one-dimensional map fitted to the observed data.

```{r, echo = FALSE}
amrc_plot_histogram_with_reference(
  values = one_dimensional_noise$stress,
  xlab = "Stress",
  fill = "#E41A1C",
  bins = 25,
  mean_line = TRUE,
  reference_line = one_dimensional_noise$observed_fit$stress
) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16)
  )
```

```{r, echo = FALSE}
one_dimensional_noise$summary %>%
  mutate(across(everything(), round, 3))
```

### Visualising one- and two-dimensional maps together

This final section compares the observed one-dimensional and two-dimensional phenotype maps directly. The one-dimensional solution is padded with a second zero axis, aligned to the 2D map by Procrustes analysis, and then plotted over the 2D configuration.

```{r, echo = FALSE}
projection_plot <- amrc_plot_one_vs_two_dimensional_projection(
  one_vs_two_dimensional$unique_phenotypes
)

projection_plot
```

```{r, echo = FALSE}
amrc_plot_histogram_with_reference(
  values = one_vs_two_dimensional$unique_phenotypes$spp_1D,
  xlab = "Stress per phenotype in 1D (%)",
  fill = "#E41A1C",
  bins = 25,
  mean_line = TRUE
)
```

```{r, echo = FALSE}
amrc_plot_histogram_with_reference(
  values = one_vs_two_dimensional$unique_phenotypes$dist_phen,
  xlab = "Distance to position on 1D map",
  fill = "#E41A1C",
  bins = 25,
  mean_line = TRUE
) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16)
  )
```
